Modeling the impact of IT- and AI-related emissions in a university setting

### 2025-08-03 doc/paper.tex
Many changes, 3rd party cloud, validation section in results

@@

@@

UZH at present does not make significant use of commercial cloud compute, but the model allows simulating the shift of some fraction of the server load to cloud VMs.

Due to lack of data, we do not attempt to model emissions from application-level cloud services such as Microsoft 365 or Google Workspace. This omission leads to undercounting total IT energy use but to overestimating the potential relative impact of AI.

UZH faculties do make use of shared off-campus supercomputing facilities, primarily the Swiss National Supercomputing Center, CSCS~\cite{cscs}. We estimate the operational impact of these workloads, but, again due to lack of data, we do not include embodied emissions. Again, this undercounts total university IT energy use but overestimates the potential relative impact of AI.

@@

We estimate inference-time emissions of current large language models based on the literature. We also include training-time emissions and embodied emissions of language models (thouse these are harder to estimate and we err on the side of overestimating them).

@@

\begin{center}
  \begin{tabular}{|l|c|c|}
  \hline
  \textbf{Device or service} & \textbf{Operation} & \textbf{Lifecycle} \\ \hline
  Student laptops \& phones & yes & yes \\
  University workstations, access points, etc. & yes & yes \\
  Cloud servers & yes & yes \\
  Cloud services & no & no \\
  Supercomputing facilities & yes & no \\
  LLMs & yes & yes \\ \hline
  \end{tabular}
  \label{tab:scopes}
  \end{center}

@@

Given all this, for each device type we can straightforwardly obtain an annual energy use (kWh) and annual embodied emissions (embodied emissions amortized over useful lifetime) (\kgcoe).

We then estimate the local grid's emissions intensity (\gcoekwh) to convert the energy use to emissions and obtain a combined total annual emissions number. We obtain emissions intensity data from Electricity Maps~\cite{electricitymaps}.

In addition to this simple linear mode, the model can also operate stochastically, instantiating devices from a distribution. However, we have insufficient data on the precise form of such a distribution, so we limit our analysis to the linear mode.

@@

We attempt to model a hypothetical shift of some workloads from servers in an on-premise data center to a commercial cloud. The parameters under consideration include:

\begin{itemize}
  \item On-premise power usage effectiveness (PUE), the ratio of total data center power (including cooling) to power used for computation.
  \item Cloud data center PUE and grid emissions intensity (since the cloud data center may be located in a different region than the on-premise data center).
  \item An average conversion ratio of on-premise servers to cloud servers.
\end{itemize}

We measure the impact of cloud servers in terms of annual operational emissions and amortized emissions due to manufacturing and disposal, in units of \kgcoe\ / year.

@@

\subsection{Parameters and assumptions}

The model's outputs are affected by several factors, and where we were unable to find good data we made assumptions that we considered reasonable. All of these parameters and assumptions are parameterizable by users of the model. We list these sources of uncertainty below.

\begin{description}
  \item[Cloud data center PUE] As of 2025, industry average power usage effectiveness is 1.56, with Google averaging 1.09 across its fleet~\cite{google:datacenter:efficiency} and Microsoft ranging from 1.12 to 1.30~\cite{microsoft:datacenter:efficiency}. We choose a PUE of 1.2 as representative of a state-of-the-art hyperscaler data center.
  \item[On-premise data center PUE] Based on limited UZH historical power data, we assume an on-premise PUE of 2.0, attributable to inefficient cooling.
  \item[Device duty cycle] We make unsupported but plausible assumptions about the number of days that different devices are used (\eg, end-user devices are not used every day) and their duty cycle throughut the day.
  \item[Electricity emissions intensity for on-premise data centers] For Switzerland, we can compute average hourly emissions intensity over 2024 based on data from Electricity Maps~\cite{electricitymaps}. Emissions vary from a high of 64 \gcoekwh\ in the middle of the night to a low of 45 \gcoekwh\ in the middle of the day. This variation is sufficiently large that we refine the model's device duty cycle to four six-hour periods over the course of the day. It allows us to specify, for example, that user-devices are mostly used during the day whereas servers run around the clock.
  \item[Electricity emissions intensity for cloud data centers and LLMs] These facilities are unlikely to be located in Switzerland, and Switzerland's grid is cleaner than average. We use the European Union's average emissions intensity of electricity generation in 2023, 207 \gcoekwh~\cite{eea:emissions:2025}.
  \item[LLM usage] We must make assumptions about the fraction of the university population that use LLMs and the frequency of use (queries / day).
\end{description}

@@

The previous section describes our overall modeling approach, which we believe is broadly applicable to university and other large campus settings. This section presents the reasoning behind specific default values we employ in the UZH evaluation.

@@

Traditional desktop computers appear to be infrequent on campus. By default we set their number to zero in our model, but we still provide parameters for them.

@@

The Swiss electricity mix is substantially cleaner than the European or North American average, so generic manufacturer estimates of lifetime usage-related emissions are unsuitable. Unfortunately, in manufacturer literature we found surprisingly few sources of data on device power consumption, as opposed to estimated lifetime CO$_2$ emissions. Based on some manufacturer data sheets, primarily from Apple, Cisco, and Dell, we model power draw of different devices while under load using the following initial parameters:

@@

Our assumptions about device duty cycle can be summarized as follows: (1) end-user devices are used during the day on weekdays, (2) servers run around-the-clock but at 80\% of peak power load, (3) mobile phones draw no power (they are unplugged) except for a few hours in the evening to charge, (4) printers are idle except for a few minutes per day.

Standby power for all devices is negligible ($\sim 1$ W), except for printers which often draw 50--100 W.

@@

\begin{center}
\begin{tabular}{|l|r|l|}
\hline
& \textbf{Embod. emiss.} & \\
\textbf{Device type} & \textbf{(\kgcoe)} & \textbf{Source} \\ \hline
Laptop & 121 & \textcite{ecoinvent}  \\ \hline
Laptop & 104 & \textcite{teehan2013} (low) \\ \hline
Laptop & 338 & \textcite{teehan2013} (high) \\ \hline
Laptop & 244 & \textcite{rarecoil} \\ \hline
Laptop & 202 & \textcite{unctadder2024} \\ \hline
Desktop & 238 & \textcite{ecoinvent} \\ \hline
Desktop & 303 & \textcite{teehan2013} \\ \hline
Desktop & 403 & \textcite{unctadder2024} \\ \hline
Desktop & 289 & \textcite{dellpcf} \\ \hline
Desktop & 277 & \textcite{boavizta} \\ \hline
Server & 383 & \textcite{teehan2013} \\ \hline
Server & 1252 & \textcite{davy2021} (Dell) \\ \hline
Server & 1912 & \textcite{davy2021} (EC2) \\ \hline
Server & 899 & \textcite{boavizta} (medium server)\\ \hline
Phone & 68 & \textcite{googlepixel8} (Pixel 8)\\ \hline
Phone & 54 & \textcite{appleiphone13} (iPhone 13)\\ \hline
Phone & 50 & \textcite{unctadder2024} \\ \hline
Phone & 50 & \textcite{lovehagen2023} \\ \hline
\end{tabular}
\label{tab:embodied_emissions}
\end{center}

@@

In the case of server, \textcite{davy2021} reports data for many Dell servers and estimates for different types of EC2 instances. The former have mean supply chain emissions of 1252 \kgcoe\ with $\sigma = 330$ kg, the latter 1912 kg with $\sigma = 885$ kg. \textcite{boavizta} reports mean supply chain emissions of 899 \kgcoe\ for a medium-sized server. We model server supply chain emissions at 1252 \kgcoe\ / device.

The range of values for mobile phones is much smaller. We model phone embodied emissions at 50 \kgcoe\ / device.

@@

\subsection{LLMs}

We instantiate as many devices as specified in the UZH spreadsheet, plus enough laptops and phones
to cover the rest of the university population.

For each device instance, we pick embodied emissions, power draw, and duty cycles either
deterministically or by sampling from the appropriate distribution.

Combining power draw and duty cycle estimates with ground-truth data about the Swiss electricity mix,
we estimate annual operational emissions and add them to the embodied emissions scaled by the device
type's expected lifetime.

For on-campus data centers, for which presumably there are electricity bills or other measures of
overall energy use, we would like to use that data to evaluate our model.

A possible future refinement will be to model the duty cycle as a function of time of day
and incorporate finer-grained electricity mix data.

\subsection{Cloud emissions}

We model cloud emissions based on the assumptions and data described in
Section~\ref{sec:cloud_data_centers}. We suggest working with UZH's cloud vendor to obtain
more accurate data.

\subsection{LLM emissions}

@@

\subsection{Cloud data centers}
\label{sec:cloud_data_centers}

The secretiveness of major cloud providers makes it difficult to estimate the embodied and operational emissions of cloud
infrastructure. The best resource we have been able to find is the Datavizta API (\cite{boavizta}).

Lacking information about UZH's workloads, we model a hypothetical UZH cloud footprint as follows:
\begin{itemize}
    \item One-for-one replacement of one campus server with one Azure D8S\_v3 instance (2x Intel
      Xeon, 8 cores, 32 GB RAM, 2TB SSD) (\cite{msftvms}).
    \item Deployment to Microsoft Azure's ``\href{https://datacenters.microsoft.com/globe/explore?info=region_switzerlandnorth}{Switzerland North}'' region .
    \item Estimation of annual emissions via Datavizta, assuming constant 50\% load: 31 \kgcoe\ for usage,
      97 \kgcoe\ for manufacturing.
\end{itemize}

Energy consumption of LLMs is being studied extensively (\cite{budennyy2022eco2ai},
\cite{castano2023exploring},
\cite{devries2023growing},
\cite{gowda2024watt},
\cite{harding2024watts},
\cite{heguerte2023estimate},
\cite{luccioni2022estimating},
\cite{luccioni2023counting},
\cite{patterson2021carbon},
\cite{rodriguez2024evaluating},
\cite{tripp2024measuring}).

@@

>>>>

% boavizta mention immaturity of assessments and variability of estimates

@@

For servers, which are mostly located in on-premise data centers, we also attempt to take into account emissions from cooling.

@@

UZH at present does not make significant use of commercial cloud compute, but the model allows simulating the shift of some fraction of the server load to cloud VMs. For this hypothetical cloud footprint, like for on-premise data centers, we attempt to model operational emissions, embodied emissions, and emissions from cooling.

UZH faculties do use shared off-campus supercomputing facilities, primarily the Swiss National Supercomputing Center, CSCS~\cite{cscs}. We estimate the operational impact of these workloads, but, due to lack of data, we do not include embodied emissions or cooling. This omission undercounts total university IT energy use but overestimates the potential relative impact of AI.

Due to lack of data, we do not model emissions from application-level cloud services such as Microsoft 365 or Google Workspace. Again, this undercounts total IT energy use but overestimate the relative impact of AI.

@@

We estimate inference-time emissions of current large language models based on the literature. We also include training-time emissions, embodied emissions, and likely cooling overhead of language models (these are harder to measure and we err on the side of overestimating them).

@@

\begin{center}
  \begin{tabular}{|l|c|c|c|}
    \hline
    \textbf{Device or service} & \textbf{Operation} & \textbf{Lifecycle} & \textbf{Cooling} \\ \hline
    Student laptops \& phones & yes & yes & n/a \\
    University workstations, access points, etc. & yes & yes & n/a \\
    University on-premise servers & yes & yes & yes \\
    Cloud servers & yes & yes & yes \\
    Cloud services & no & no & no \\
    Supercomputing facilities & yes & no & no \\
    LLMs & yes & yes & yes \\ \hline
  \end{tabular}
  \label{tab:scopes}
\end{center}

@@

Given all this, for each device type we can straightforwardly obtain an annual energy use (kWh) and annual embodied emissions (embodied emissions amortized over useful lifetime) (\kgcoe). We then estimate the local grid's emissions intensity (\gcoekwh) to convert the energy use to emissions and obtain a combined total annual emissions number. (In addition to this simple linear mode, the model can also operate stochastically, instantiating devices from a distribution. However, we have insufficient data on the precise form of such a distribution, so we limit our analysis to the linear mode.)

Note that for Switzerland we can compute average hourly emissions intensity over 2024 based on data from Electricity Maps~\cite{electricitymaps}. Emissions vary from a high of 64 \gcoekwh\ in the middle of the night to a low of 45 \gcoekwh\ in the middle of the day. This variation is sufficiently large that we have refined the model's device duty cycle to represent four six-hour periods over the course of the day. It allows us to specify, for example, that user devices are mostly used during the day whereas servers run around the clock.

For servers (in on-premise data centers) we also account for cooling overhead. Based on limited UZH historical power data, we estimate that the on-premise power usage effectiveness (PUE), the ratio of total data center power (including cooling) to power used for computation, is approxmiately 2. This value, higher than the commercial average of 1.56~\cite{google:datacenter:efficiency}, is attributable to inefficient cooling.

@@

We attempt to model a hypothetical shift of some workloads from servers in an on-premise data center to a commercial cloud. We model the shift as a fraction of servers that are moved to the cloud, the average conversion ratio of on-premise servers to cloud servers, and cloud data center PUE and electricity emissions intensity (since the cloud data center may be located in a different region than the on-premise data center).

@@

@@

The previous section describes our overall modeling approach, which we believe is broadly applicable to university and other large campus settings. This section presents the reasoning behind specific values we employ in the UZH evaluation.

@@

Traditional desktop computers appear to be infrequent on campus. By default we set their number to zero in our model, but we still provide the framework to model them if desired.

@@

The Swiss electricity mix is substantially cleaner than the European or North American average, so generic manufacturer estimates of lifetime usage-related emissions are unsuitable. Unfortunately, in manufacturer documentation we found surprisingly few sources of data on device power consumption, as opposed to estimated lifetime CO$_2$ emissions. Based on some manufacturer data sheets, primarily from Apple, Cisco, and Dell, we model power draw of different devices while under load using the following initial parameters:

@@

Our assumptions about device duty cycle can be summarized as follows: (1) end-user devices are used during the day on weekdays, (2) servers run around-the-clock but at 80\% of peak power load, (3) mobile phones draw no power (they are unplugged) except for a few hours in the evening to charge, (4) printers are idle except for a few minutes per day. Standby power for all devices is negligible ($\sim 1$ W), except for printers which often draw 50--100 W.

@@

\begin{center}
  \begin{tabular}{|l|r|l|}
    \hline
    & \textbf{Embod. emiss.} & \\
    \textbf{Device type} & \textbf{(\kgcoe)} & \textbf{Source} \\ \hline
    Laptop & 121 & \textcite{ecoinvent}  \\ \hline
    Laptop & 104 & \textcite{teehan2013} (low) \\ \hline
    Laptop & 338 & \textcite{teehan2013} (high) \\ \hline
    Laptop & 244 & \textcite{rarecoil} \\ \hline
    Laptop & 202 & \textcite{unctadder2024} \\ \hline
    Desktop & 238 & \textcite{ecoinvent} \\ \hline
    Desktop & 303 & \textcite{teehan2013} \\ \hline
    Desktop & 403 & \textcite{unctadder2024} \\ \hline
    Desktop & 289 & \textcite{dellpcf} \\ \hline
    Desktop & 277 & \textcite{boavizta:api} \\ \hline
    Server & 383 & \textcite{teehan2013} \\ \hline
    Server & 1252 & \textcite{davy2021} (Dell) \\ \hline
    Server & 1912 & \textcite{davy2021} (EC2) \\ \hline
    Server & 899 & \textcite{boavizta:api} (medium server)\\ \hline
    Phone & 68 & \textcite{googlepixel8} (Pixel 8)\\ \hline
    Phone & 54 & \textcite{appleiphone13} (iPhone 13)\\ \hline
    Phone & 50 & \textcite{unctadder2024} \\ \hline
    Phone & 50 & \textcite{lovehagen2023} \\ \hline
    \end{tabular}
  \label{tab:embodied_emissions}
\end{center}

@@

In the case of servers, \textcite{davy2021} reports manufacturer data for many Dell servers and estimates for different types of EC2 instances. The former have mean supply chain emissions of 1252 \kgcoe\ with $\sigma = 330$ kg, the latter 1912 kg with $\sigma = 885$ kg. \textcite{boavizta:api} reports mean supply chain emissions of 899 \kgcoe\ for a medium-sized server. We model server supply chain emissions at 1252 \kgcoe\ / device.

The variability of estimates for mobile phones is much smaller. We model phone embodied emissions at 50 \kgcoe\ / device.

@@

The secretiveness of major cloud providers makes it difficult to estimate the embodied and operational emissions of cloud infrastructure. The best resource we have found is the Datavizta API by the Boavizta working group~\cite{boavizta:api}.

Lacking information about UZH's workloads, we model a hypothetical UZH cloud footprint as the replacement of one campus server with a configurable number (by default, one) of Azure {\tt standard\_d48s\_v3} instances (Intel Xeon, 48 cores, 192 GB RAM, 2TB SSD)~\cite{msftvms}. Based on data from Boavizta, such a machine has embodied emissions of 667 \kgcoe\ and mean electricity power usage of 255 W. We assume deployment to Microsoft Azure's ``\href{https://datacenters.microsoft.com/globe/explore?info=region_switzerlandnorth}{Switzerland North}'' region, which makes sense for a Swiss university and means that the electricity mix is the same as for on-premise facilities. We estimate a power usage effectiveness of 1.15 for this facility. (As of 2025, average PUE for cloud computing is 1.56, with Google averaging 1.09 across its fleet~\cite{google:datacenter:efficiency} and Microsoft ranging from 1.14 to 1.19 in Europe~\cite{microsoft:datacenter:efficiency}.)

Given all these parameters, we can then compute cloud server operational emissions, embodied emissions, and cooling overhead as we do for on-premise servers.

We must make assumptions about the fraction of the university population that use LLMs and the frequency of use (queries / day).

@@

Energy consumption of LLMs is being studied extensively~\cite{budennyy2022eco2ai,castano2023exploring,devries2023growing,gowda2024watt,harding2024watts,heguerte2023estimate,luccioni2022estimating,luccioni2023counting,patterson2021carbon,rodriguez2024evaluating,tripp2024measuring}.

@@

Electricity emissions intensity for cloud data centers and LLMs] These facilities are unlikely to be located in Switzerland, and Switzerland's grid is cleaner than average. We use the European Union's average emissions intensity of electricity generation in 2023, 207 \gcoekwh~\cite{eea:emissions:2025}.
